Comparing summed probability distributions of shoreline and radiocarbon dates from the Mesolithic Skagerrak coast of Norway

### 2023-07-03 analysis/paper/paper.Rmd
Edits from Steinar

---
title: "Comparing summed probability distributions of shoreline and radiocarbon dates from the Mesolithic Skagerrak coast of Norway"
# author: 
# - 'Isak Roalkvam'
# - 'University of Oslo, Institute of Archaeology, Conservation and History'
  # - IsakRoalkvam:
  #     email: isak.roalkvam@iakh.uio.no
  #     institute: [iakh]
      # correspondence: true
#   - name: Steinar Solheim
#     email: steinar.solheim@khm.uio.no
#     institute: [mch]
#     correspondence: false
institute:
  - iakh: Institute of Archaeology, Conservation and History, University of Oslo
#   - mch: Museum of Cultural History, University of Oslo
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::pdf_document2:
      fig_caption: yes
      toc: false
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      # - --lua-filter=../templates/scholarly-metadata.lua
      # - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
header-includes:
  \usepackage{placeins}
# always_allow_html: true
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
# abstract: |
#   Text of abstract
# keywords: |
#   keyword 1; keyword 2; keyword 3
# highlights: |
#   These are the highlights. 
---

@@

# Introduction
Population size is regarded as an important driver of cultural variation, and is of critical importance to our understanding of past human societies [e.g. @french2021; @shennan2000]. The frequency distribution of radiocarbon dates has been used extensively as one proxy for past relative population sizes [e.g. @crema2022; @french2021], including in Norwegian Stone Age archaeology [@solheim2018; @solheim2020a; @nielsen2019; @nielsen2021; @bergsvik2021; @jorgensen2020; @lundstrom2021; see also @persson2009]. The potentially immense value of insights into past population dynamics, combined with the ubiquity of radiocarbon dates and the relative ease with which these can now be processed within what has been termed a dates-as-data methodology [@rick1987] has undoubtedly contributed to the popularity of the approach. Several limitations and forms of criticism have, however, been directed at these procedures. Some of the objections are of a methodological nature, while others pertain to the underlying logic and the degree to which there is likely to be a direct connection between the frequency of ^14^C-dates and population dynamics [e.g. @carleton2021; @torfing2015; @ward2021]. What appears to be agreed upon by practitioners and critics alike is that radiocarbon dates are best analysed in this manner when compared and contrasted to other proxies for past population dynamics and other variables that might impact these [@french2016; @palmisano2017; @palmisano2021; @bergsvik2021]. This paper reports on and begins to unpack the relationship between two measures that have been linked to relative population size in the context of the Mesolithic Skagerrak coast in south-eastern Norway, namely summed probability distribution of calibrated radiocarbon dates and the summed probability distribution of shoreline dated sites (hereafter RSPD and SSPD, respectively).

# Background
Large parts of the post-glacial landscape of Northern Scandinavia is characterised by dramatic isostatic uplift that has led to a net sea-level fall throughout the Holocene, despite eustatic sea-level rise [e.g. @mörner1979]. As coastal foragers appear to have predominantly settled on or close to the contemporaneous shoreline, this can be utilised to assign an approximate date to the sites. This is done by coupling the present-day altitude of the sites with reconstructions of past shoreline displacement -- a method known as shoreline dating. This is not least useful for dating the large number of surveyed sites in the region where other temporal data that often follow with an excavation are not available, such as radiocarbon dates or typological indicators in artefact inventories.

The frequency of shoreline dated sites has also been compared to RSPDs in the past [@solheim2018; @tallavaara2020]. However, this has been done by finding point estimates of shoreline dates that are then aggregated in somewhat arbitrary bins of 200 or 500 years. This therefore does not take into account the uncertainty in the distance between the sites and the contemporaneous shoreline, nor the impact the variability in the rate of sea-level change has on the precision of the dates that can be achieved with the method. In a recent study, @roalkvam2023 has presented a probabilistic method for shoreline dating that takes these parameters into account, thus setting the stage for a more refined investigation of the relationship between frequency of ^14^C-dates and shoreline dated sites. The parametrisation of the method was based on simulating the distance between sites with ^14^C-dates and the prehistoric shoreline along the Skagerrak coast in the region between Horten municipality in the northeast to Arendal municipality in the southwest (see map in Figure \@ref(fig:map)). The results of the analysis indicates that the sites tend to have been located close to the shoreline until just after 4000 BCE when a few sites become more withdrawn from the shoreline, followed by a clear break around 2500 BCE, at which point shoreline dating appears to loose its utility. Thus, this geographical and temporal limit are also used for this study.

This paper is decidedly exploratory. While some speculative thoughts concerning the relationship between the variables outlined above underlie the analysis, these could not be instantiated as concrete hypotheses. The data have thus been explored using a set of standard models to begin to unpick patterns and grasp the relationship between the variables. Furthermore, the dates as data approach is dependent on there being a direct link between the past generation of material that ultimately become ^14^C samples, and population size. The sum of shoreline dated sites, on the other hand, is determined by site frequency, and if used as a proxy for population size is therefore dependent on there being a connection between site count and population size. If a comparison of these proxies do not, or only partially correspond, it is thus an open question what factors have impacted either distribution to cause this discrepancy, and which measure, if any, reflects true population dynamics. The issue will therefore initially demand an open and exploratory approach where a multitude of explanatory and confounding effects can be drawn on to suggest explanations of any observed pattern.

@@

To what degree the radiocarbon record is determined by past population numbers might vary both geographically and chronologically, based on variation in investigatory and taphonomic factors [@surovell2009b; @bluhm2019], as well as cultural processes within prehistoric populations. One example of the latter is the difference that might exist between farmer and forager populations, where @freeman2018 have suggested that an increased per capita energy consumption introduced with farming means that ^14^C-dates should not be weighted equally when making relative population estimates across such populations. Similarly, while site counts have also been invoked for the analysis of past population dynamics, these are likely to be impacted by factors such as land-use and mobility patterns, and settlement nucleation and dispersion [@palmisano2017]. However, these could be considered theoretical issues that have implications for how fluctuations in these proxies should be interpreted, and possibly weighted differently according some criteria. Before any such fluctuations are given any substantive interpretation, however, there are a host of methodological issues that have to be considered.

The most critical of these follow from the fact that the summation of the probabilities associated with the dates for the SPDs is not a statistically coherent procedure. This is because the summed probabilities can no longer been seen as probabilities, but rather represent the combination of events and uncertainties, making the two indistinguishable, and rendering the interpretation of the resulting sum difficult [@blackwell2003; @crema2022]. As @timpson2021[2] put it: 'the SPD is not *the* single best explanation of the data, nor even *a* explanation of the data, but rather a conflation of many possible explanations simultaneously, each of which is mired by the artefacts inherited from the calibration wiggles.' The SPD is not a model. It is the combined representation of a range of possible explanations for the data -- the frequency of dated events combined with the variable uncertainty associated with these [@carleton2021]. This means that an SPD cannot be directly analysed to draw inferences on population dynamics, nor can it be directly compared to other time-series data [@carleton2018; @timpson2021]. While this problem can never be entirely resolved, a range of approaches have been developed in an attempt to work around this issue.

The most commonly applied of these is a null-hypothesis significance testing approach by means of Monte Carlo simulation, as introduced by @shennan2013 and later expanded upon by @timpson2014. This works by comparing the observed RSPD with a series of simulated RSPDs, generated from a null-model. These null-models are typically a uniform, exponential or logistic distribution. These are chosen *a priori* with reference to common long-term population dynamics and are typically parametrised by fitting the models by means of regression or maximum likelihood estimation (MLE). The result from these simulations are then used to create a 95% critical envelope representing the null model. The proportion of the observed RSPD that falls outside this envelope is then used to estimate a global p-value indicating whether the null model can be rejected. In the case that it can, the portions of the observed RSPD that falls outside this envelope can subsequently be interpreted as representing potentially meaningful demographic events, relative to the null model. However, care has to be taken in how these are interpreted. First, this follows from the fact that 5% of the deviations from the critical envelope can be expected to be random, and there is no way to know which deviations this pertains to. Secondly, a for example exponential null model fit to the data is only one of an infinite set of exponential models with different growth rates that could be used. While a model fit by means of MLE will likely have a reasonable growth rate, and by extension exclude many other exponential fits as likely to explain the data, this can be difficult to determine. Finally, the p-value only indicates whether or not the null-model as a whole can be rejected as an explanation of the data, and does not provide statistical justification for interpreting local deviations themselves as meaningful demographic signals, as has often been the case [see review by @crema2022; @timpson2021].

## Model comparison and continous piece-wise linear models

The procedure outlined above represent the standard approach in efforts to model population dynamics based on RSPDs. This can be helpful for determining whether the population behaviour has followed the trajectories of standard theoretical models, and is a useful first step here also for the treatment of shoreline dated sites, as few assumptions concerning the verall development of this proxy could be made. However, standard models frequently used to evaluate RSPDs are consistently being rejected in the literature, and deviations from these rejected null-models do not in themselves allow for any interpretations of the deviations, nor a subsequent direct analysis of the SPDs themselves. As a result, and by improving on @goldstein2016, @timpson2021 have recently suggested an alternative framework for modelling past population dynamics by use of RSPDs, published with the accompanying R package  *ADMUR*. This involves fitting continuous piece-wise linear (CPL) models to the data, in addition to the standard models, subjecting these to a procedure of model comparison, and finally evaluating the best model using a goodness-of-fit test, in line with the Monte Carlo approach outlined above. The CPL models are defined by discrete linear periods of development between hinge points that mark points in time in which the trajectory of the population proxy changes direction. This therefore allows for the identification of key events and a conservative description of population dynamics between these. The approach thus also provides a clearer inferential justification for a further analysis and substantive interpretation of the identified best model. Following the evaluation of standard models as outlined above, the data were therefore subsequently subjected to the approach presented by @timpson2021.

@@

```{r map, echo = FALSE, fig.cap = "Map of the study area and shoreline dated sites. Black lines indicate the borders between municipalities. The surveyed sites included in the analysis are the ones given a quality score of 3 or higher using the framework in Table 1.", out.width = "500px", fig.align="center", regfloat = TRUE}

@@

To account for investigatory bias that can result from variable sampling intensity between sites, the summing of radiocarbon dates with *ADMUR* starts with binning dates from the same sites that fall within 200 uncalibrated ^14^C years of each other when measured from the mean ^14^C age. All dates are then calibrated using a resoltuion of 5 years and then normalised to sum to unity. The RSPDs for each site-phase are then summed to achieve the final SPD, which is in turn normalised. All calibrations were performed using the IntCal20 calibration curve [@reimer2020].

The final RSPD was subjected to the standard null-hypothesis testing approach through Monte Carlo simulation, as introduced above [see @shennan2013; @timpson2014], by fitting an exponential, logistic and uniform model to the observed RSPD. Following @timpson2021, the model parameters were identified by MLE search, using the differential evolution optimization algorithm *DEoptimR* [@brest2006]. The Monte Carlo simulations were then performed using the *SPDsimulationTest()* function from *ADMUR*. For each model, a series of individual calendar years are drawn from the model distribution, with replacement, the number of which equals the number of bins in the observed RSPD. These are then 'uncalibrated' to a single value on the ^14^C scale and a random error from among the observed errors is added to the date. These are then calibrated back to the calendar scale and finally summed. Here, this procedure was repeated 10000 times for each null model. The 2.5th and 97.5th quantile of the resulting summed probability for each year across all simulations are then retrieved to create the 95% critical envelope with which to compare the observed RSPD. The degree to which the observed RSPD deviates from the critical envelope is then used to calculate a global p-value, indicating whether or not the null can be rejected.

@@

Four geological reconstructions of shoreline displacement in the region lays the foundation for the method as implemented here (Figure \@ref(fig:shoredate)B). These shoreline displacement curves are from Horten [@romundset2021], Porsgrunn [@sørensen2014; @sørensen2014b; @sorensen2023], Tvedestrand [@romundset2018; @romundset2018b] and Arendal [@romundset2018b], each associated with a shoreline isobase along which the trajectory for relative sea-level change has been the same [see @svendsen1987]. The first step in the dating procedure is to interpolate the shoreline displacement to the location to be dated. This is done by inverse distance weighting [e.g. @conolly2020], interpolating the relative sea-level change to the location of a site to be dated, based on it's distance from the isobases of the displacement curves.

@@

With an estimated trajectory of shoreline displacement at the site to be dated, the elevation of the site can be drawn on to find a likely date for when it was in use, under the assumption that it was in use when located on or close to the shoreline. Figure \@ref(fig:shoredate)C indicates where the elevation of the example site intersects the interpolated displacement curve. As it can be reasonably be assumed that a site was not in use when located under water, this effectively represents a *terminus post quem* (TPQ) date. However, a fairly common approach to shoreline dating has been find where the elevation of the site intersects the mean of the upper and lower limit of the displacement curve, and then adding a somewhat arbitrary error of $\pm$ 100 years to this point-estimate. However, finding TPQ dates that do not account for the likely distance between site and sea when the site was in use limits the further inferential steps that can be taken. Furthermore, the more common practice of adding a constant error to a point-estimate does not take into account that the rate of shoreline displacement varies both spatially and temporally, and therefore introduces bias to the result on both the spatial and temporal scale, while also overestimating the precision of the method in most cases.

@@

The analysis in @roalkvam2023 found that site phases ^14^C-dated to before 2500 BCE were found to have a likely elevation above sea-level that can be reasonably approximated by a gamma distribution with shape =  0.286 and scale = 20.833. This is incorporated in the procedure for shoreline dating used for this study (Figure \@ref(fig:shoredate)D). When performing this dating procedure, the gamma distribution is sequentially stepped through and transferred to the calendar scale by uniformly distributing the probability across the years in the range between the lower and upper limit of the interpolated displacement curve. This procedure thus gives the the shoreline date of a site when accounting for the likely elevation of the site above sea-level when it was in use [see @roalkvam2023 for details]. Given that the shoreline displacement curves have no inversions and should therefore before commutative [cf. @weninger2015, 545], each shoreline date is normalised to sum to unity. To reduce the computational cost of the simulation procedures to follow below, the gamma distribution is here stepped through at increments of 0.1 m and the calendar scale is kept at a resolution of 5 years.

@@

The implications of different approaches to shoreline dating and summing the results is further illustrated in Figure \@ref(fig:sumshore). In Figure \@ref(fig:sumshore)A  the sites to be shoreline dated are plotted according to their elevation. Due to the variable rates of shoreline displacement within the study region, the same elevation does not necessarily equate to the same date, and so this does not directly tell us much about their temporal distribution. For example, as the trajectory for sea-level regression is starker towards the north-east, the same elevation at a location to the south-west implies a younger date that one towards the north-east.

When the frequency distribution of shoreline dates have been investigated in the past, this has typically involved finding point-estimates of dates by either using the mean of displacement curves, or by taking mean of the TPQ dates for each site (as illustrated in Figure \@ref(fig:shoredate)C). These point-estimates are then aggregated in bins of 200 or 500 years [e.g. @breivik2014; @solheim2018; @fossum2020; @mjærum2022; @tallavaara2020]. This approach is illustrated in Figure \@ref(fig:sumshore)B. The SSPD resulting from instead using the probabilistic method of shoreline dating, given in Figure \@ref(fig:sumshore)C, demonstrates that accounting for the site-sea relationship gives a substantially different result compared to the more traditional approach, which is clearly indicated by the shifts in the positions of peaks and troughs in the distribution after c. 8000 BCE.

@@

However, in the same way is with RSPDs, the SSPD cannot be directly interpreted as reflecting the intensity of sites over time, due to the summation issues introduced above. Furthermore, similar to how characteristics of the calibration curve can introduce bias to the RSPD, the same is true for the SSPD, where the local trajectory of relative sea-level change effectively functions as the calibration curve for each site. As the shoreline displacement curves are interpolated to the sites based on their location along a south-west--north-east gradient, each site is effectively associated with a unique shoreline displacement curve, provided they are not located on exactly the same isobase. With analogy to the radiocarbon methodology, this would be equivalent to each date being associated with a unique calibration curve. As it would be computationally prohibitive to interpolate the shoreline displacement trajectory for each date to be simulated in the Monte Carlo procedure, one shoreline displacement curve was initially interpolated to the centre of each of a series of 2 km wide line segments running perpendicular to the shoreline gradient between the extremes of the distribution of sites. These intervals were then assigned a weight based on how the density of observed sites is distributed among them (Figure \@ref(fig:incpolys)).

@@

```{r spdmc, echo = FALSE, fig.cap = "Monte Carlo simulation for shoreline and radiocarbon dates. A) Summed probability of shoreline dated sites (n = 921) compared to an exponential null model. B) SSPD compared to a logistic null model. C) SSPD compared to a uniform null model. D) Summed probability of calibrated radiocarbon dates (n = 310, bins = 134) compared to an exponential null model. E) RSPD compared to a logistic null model. F) RSPD compared to a uniform null model.", out.width = "500px", fig.align="center", regfloat = TRUE}
knitr::include_graphics(here("analysis/figures/mc.png"))
```
For the RSPD, the logistic model achieved a p-value of 0.108, and could therefore not be rejected. Based on the available data, this therefore represents the best explanation of the data. For the SSPD all of the standard models could be rejected.

<!-- We can also add that accepting this model means that interpreting any deviations from the Monte Carlo enevelope is inappropriate, although this has sometimes been the case. -->

>>>>

---
title: "Comparing summed probability distributions of shoreline and radiocarbon dates from the Mesolithic Skagerrak coast of Norway"
# author:
#   - Isak Roalkvam, University of Oslo, Institute of Archaeology, Conservation and History
#   - Steinar Solheim, University of Oslo, Museum of Cultural History
author:
  - Isak Roalkvam:
      email: isak.roalkvam@iakh.uio.no
      institute: [iakh]
      correspondence: true
  - name: Steinar Solheim
    email: steinar.solheim@khm.uio.no
    institute: [mch]
    correspondence: false
institute:
  - iakh: University of Oslo, Institute of Archaeology, Conservation and History
  - mch: University of Oslo, Museum of Cultural History
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::pdf_document2:
      fig_caption: yes
      toc: false
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - '--lua-filter=../templates/scholarly-metadata.lua'
      - '--lua-filter=../templates/author-info-blocks.lua'
      - '--lua-filter=../templates/pagebreak.lua'
header-includes:
  \usepackage{placeins}
# always_allow_html: true
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
# abstract: |
#   Text of abstract
# keywords: |
#   keyword 1; keyword 2; keyword 3
# highlights: |
#   These are the highlights. 
---

@@

# Introduction
Variation in population patterns are regarded important drivers and consequences of cultural variation, and of critical importance to our understanding of past human societies [e.g. @french2021; @shennan2000; @shennan2020]. During the last decade, the frequency distribution of radiocarbon dates have become the preferred proxy for archaeologists studying prehistoric population dynamics [@crema2022; @french2021; @shennan2013], owing in part to the fact that radiocarbon dates represent the most temporally and spatially widespread archaeological data available for studying past population size and density [@bird2022]. This has also been the case for Norwegian Stone Age archaeology [@solheim2018; @solheim2020a; @nielsen2019; @nielsen2021; @bergsvik2021; @jorgensen2020; @lundstrom2021; see also @persson2009].

The potentially immense value of insights into population dynamics, combined with the ubiquity of radiocarbon dates and the relative ease with which these can now be processed within what has been termed a dates-as-data methodology [@rick1987] has undoubtedly contributed to the popularity of the approach. Several limitations and forms of criticism have, however, been directed at these procedures. Some of the objections are of a methodological nature, while others pertain to the underlying logic and the degree to which there is likely to be a direct connection between the frequency of ^14^C-dates and population dynamics [e.g. @carleton2021; @torfing2015; @ward2021]. What appears to be agreed upon by practitioners and critics alike is that radiocarbon dates are best analysed in this manner when compared and contrasted to other proxies for past population dynamics and other variables that might impact these [@bevan2017; @french2021; @palmisano2017; @palmisano2021]. In this paper, we aim to report on and begin to unpack the relationship between two measures that have been linked to relative population size in the context of the Mesolithic Skagerrak coast in south-eastern Norway, namely summed probability distribution of calibrated radiocarbon dates (RSPD) and the summed probability distribution of shoreline dated sites (SSPD). We discuss how the data compare with each other as well as how it can be used to understand cultural historical development in the region through the Mesolithic, c. 9500--4000 BCE.

# Background
Following the melting of the Fennoscandian Ice Sheet, large parts of the post-glacial landscape of Northern Scandinavia has been characterised by dramatic isostatic uplift that has led to a net sea-level fall throughout the Holocene, despite eustatic sea-level rise [e.g. @mörner1979]. The majority of Mesolithic sites along the Norwegian Skagerak coast are interpreted as having been shore bound or at least located close to the contemporaneous shoreline at the time of use. As coastal foragers in this region appear to have predominantly settled on or close to the contemporaneous shoreline during early and mid-Holocene, this can be utilised to assign an approximate date to the sites. This is done by coupling the present-day altitude of the sites with reconstructions of past shoreline displacement -- a method known as shoreline dating. method has a long tradition in Norwegian archaeology [@brogger1905], and is especially useful for dating the large number of surveyed archaeological sites when temporal data that often follow with an excavation, such as radiocarbon dates or typological indicators in artefact inventories, are not available.  Previous studies shave suggested that shoreline dated sites can be used as a proxy to study temporal variation in population and settlement patterns [@solheim2018; @tallavaara2020; @jørgensen2020].

The frequency of shoreline dated sites has been compared to RSPDs in the past by finding point estimates of shoreline dates that are aggregated in somewhat arbitrary bins of 200 or 500 years. This therefore does not consider the uncertainty in the distance between the sites and the contemporaneous shoreline, nor the impact the variability in the rate of sea-level change has on the precision of the dates that can be achieved with the method. In a recent study, @roalkvam2023 has presented a probabilistic method for shoreline dating that takes these parameters into account, thus setting the stage for a more refined investigation of the relationship between frequency of ^14^C-dates and shoreline dated sites. The parametrisation of the method was based on simulating the distance between sites with ^14^C-dates and the prehistoric shoreline along the Skagerrak coast in the region between Horten municipality in the northeast to Arendal municipality in the southwest (see map in Figure \@ref(fig:map)). The results of the analysis indicates that the sites tend to have been located close to the shoreline until just after 4000 BCE when a few sites become more withdrawn from the shoreline [see also @nielsen2021a], followed by a clear break around 2500 BCE, at which point shoreline dating appears to lose its utility. Thus, this geographical and temporal limit are also used for this study.   
The present paper is decidedly exploratory. While some thoughts concerning the relationship between the variables outlined above underlie the analysis, these could not be instantiated as concrete hypotheses. Rather, we explore the data by using a set of standard models to begin to unpack patterns and grasp the relationship between the variables. Furthermore, the dates-as-data approach is dependent on there being a link between how prehistoric people generated material that become ^14^C samples, and reflect population dynamics [cf- @freeman2018]. The sum of shoreline dated sites, on the other hand, is determined by site frequency, and its use as a population proxy is therefore dependent on there being a connection between site count and population size and density. If a comparison of these proxies do not, or only partially correspond, it is thus an open question what factors have impacted either distribution to cause this discrepancy, and which measure, if any, reflects true population dynamics. The issue will therefore initially demand an open and exploratory approach where a multitude of explanatory and confounding effects can be drawn on to suggest explanations of any observed pattern.

@@

To what degree the radiocarbon record is determined by past population numbers might vary geographically and chronologically, based on variation in investigatory and taphonomic factors [@surovell2009b; @bluhm2019], as well as cultural processes within prehistoric populations. One example of the latter is the difference that might exist between farmer and forager populations, where @freeman2018 have suggested that an increased per capita energy consumption introduced with farming means that ^14^C-dates should not be weighted equally when making relative population estimates across such populations. Similarly, while site counts have also been invoked for the analysis of past population dynamics, these are likely to be impacted by factors such as land-use and mobility patterns, and settlement nucleation and dispersion [@palmisano2017], as well as variation in archaeological investigations and what has been taken to consitute a 'site'. However, these could be considered theoretical issues that have implications for how fluctuations in these proxies should be interpreted, and possibly weighted differently according some criteria. Before any such fluctuations are given any substantive interpretation, however, there is a host of methodological issues that have to be considered [see @grocutt2020; @crema2022 for a review].

The most critical of these follow from the fact that the summation of the probabilities associated with the dates for the SPDs is not a statistically coherent procedure. This is because the summed probabilities can no longer been seen as probabilities, but rather represent the combination of events and chronological uncertainties, making the two indistinguishable, and rendering the interpretation of the resulting sum difficult [@blackwell2003; @crema2022]. As @timpson2021[2] put it: 'the SPD is not *the* single best explanation of the data, nor even *a* single explanation of the data, but rather a conflation of many possible explanations simultaneously, each of which is mired by the artefacts inherited from the calibration wiggles.' The SPD is not a model. It is the combined representation of a range of possible explanations for the data -- the frequency of dated events combined with the variable uncertainty associated with these [@carleton2021]. This means that visual inspections of SPDs can be misleading, that they cannot be directly analysed to draw inferences on population dynamics, nor can they be directly compared to other time-series data [@carleton2018; @timpson2021; @crema2022]. While this problem can never be entirely resolved, a range of approaches have been developed in an attempt to work around this issue.

The most commonly applied of these is a null-hypothesis significance testing approach by means of Monte Carlo simulation, first introduced by @shennan2013 and later expanded upon by @timpson2014 and @crema2020. This works by comparing the observed RSPD with a series of simulated RSPDs, generated from a null-model. These null-models are typically a uniform, exponential or logistic distribution. These are chosen *a priori* with reference to common long-term population dynamics and are typically parametrised by fitting the models by means of regression or maximum likelihood estimation (MLE). The result from these simulations are then used to create a 95% critical envelope representing the null model. The proportion of the observed RSPD that falls outside the simulated envelope (i.e. positive or negative divergence of the observed SPD from the fitted null model), is then used to estimate a global p-value indicating whether the null model can be rejected. In the case that it can, the portions of the observed RSPD that falls outside this envelope can subsequently be interpreted as representing potentially meaningful demographic events, relative to the null model. However, care has to be taken in how these are interpreted. First, this follows from the fact that 5% of the deviations from the critical envelope can be expected to be random, and there is no way to know which deviations this pertains to. Secondly, a for example exponential null model fit to the data is only one of an infinite set of exponential models with different growth rates that could be used. While a model fit by means of MLE will likely have a reasonable growth rate, and by extension exclude many other exponential fits as likely to explain the data, this can be difficult to determine. Finally, the p-value only indicates whether or not the null-model as a whole can be rejected as an explanation of the data, and does not provide statistical justification for interpreting if local deviations themselves are meaningful demographic signals [see review by @crema2022; @timpson2021].

## Model comparison and continuous piece-wise linear models

The procedure outlined above represent the standard approach in efforts to model population dynamics based on RSPDs. This can be helpful for determining whether the population patterns has followed the trajectories of standard theoretical models. It is also a useful first step here also for the treatment of shoreline dated sites, as few assumptions concerning the overall development of this proxy could be made. However, standard models frequently used to evaluate RSPDs are consistently being rejected in the literature, and, as mentioned above, deviations from these rejected null-models do not in themselves allow for any interpretations of the local deviations, nor a subsequent direct analysis of the SPDs themselves. As a result, and by improving on @goldstein2016, @timpson2021 have recently suggested an alternative framework for modelling past population dynamics by use of RSPDs, published with the accompanying R package  *ADMUR*. This involves fitting continuous piece-wise linear (CPL) models to the data, in addition to the standard models, subjecting these to a procedure of model comparison, and finally evaluating the best model using a goodness-of-fit test, in line with the Monte Carlo approach outlined above. The CPL models are defined by discrete linear periods of development between hinge points that mark points in time in which the trajectory of the population proxy changes direction. This allows for the identification of key demographic events and a conservative description of population dynamics between these. The approach thus provides a clearer inferential justification for a further analysis and substantive interpretation of the identified best model. Following the evaluation of standard models as outlined above, the data were subsequently subjected to the approach presented by @timpson2021.

@@

```{r map, echo = FALSE, fig.cap = "A) Map displaying the location of the study area and shoreline dated sites. Black lines indicate the borders between municipalities. The surveyed sites included in the analysis are the ones given a quality score of 3 or higher using the framework in Table 1. B) The distribution of radiocarbon dates employed in the study.", out.width = "500px", fig.align="center", regfloat = TRUE}

@@

To account for investigatory bias that can result from variable sampling intensity between sites, the summing of radiocarbon dates with *ADMUR* starts with binning dates from the same sites that fall within 200 uncalibrated ^14^C years of each other when measured from the mean ^14^C age. All dates are then calibrated using a resolution of 5 years and normalised to sum to unity. The RSPDs for each site-phase are summed to achieve the final SPD, which is in turn normalised. All calibrations were performed using the IntCal20 calibration curve [@reimer2020].

The final RSPD was subjected to the standard null-hypothesis testing approach through Monte Carlo simulation, as introduced above [see @shennan2013; @timpson2014], by fitting an exponential, logistic and uniform model to the observed RSPD. Following @timpson2021, the model parameters were identified by MLE search, using the differential evolution optimization algorithm *DEoptimR* [@brest2006]. The Monte Carlo simulations were then performed using the *SPDsimulationTest()* function from *ADMUR*. For each model, a series of individual calendar years are drawn from the model distribution, with replacement, the number of which equals the number of bins in the observed RSPD. These are then 'uncalibrated' to a single value on the ^14^C scale and a random error from among the observed errors is added to the date. These are then calibrated back to the calendar scale and finally summed. Here, this procedure was repeated 10000 times for each null model. The 2.5th and 97.5th quantile of the resulting summed probability for each year across all simulations are then retrieved to create the 95% critical envelope with which to compare the observed RSPD. The degree to which the observed RSPD deviates from the critical envelope is used to calculate a global p-value, indicating whether the null model can be rejected.

@@

Four geological reconstructions of shoreline displacement in the region lays the foundation for the method as implemented here (Figure \@ref(fig:shoredate)B). These shoreline displacement curves are from Horten [@romundset2021], Porsgrunn [@sørensen2014; @sørensen2014b; @sorensen2023], Tvedestrand [@romundset2018; @romundset2018b] and Arendal [@romundset2018b], each associated with a shoreline isobase along which the trajectory for relative sea-level change has been the same [see @svendsen1987]. The first step in the dating procedure is to interpolate the shoreline displacement to the site to be dated. This is done by inverse distance weighting [e.g. @conolly2020], interpolating the relative sea-level change to the location of an archaeological site, based on its distance from the isobases of the displacement curves.

@@

With an estimated trajectory of shoreline displacement at the site to be dated, the elevation of the site can be drawn on to find a likely date for when it was in use, under the assumption that it was in use when located on or close to the shoreline. Figure \@ref(fig:shoredate)C indicates where the mean elevation of the polygon representing the example site intersects the interpolated displacement curve. As it can reasonably be assumed that a site was not in use when located under water, this effectively represents a *terminus post quem* (TPQ) date. A common approach to shoreline dating has been to find where the elevation of the site intersects the mean of the upper and lower limit of the displacement curve, and then adding a somewhat arbitrary error of $\pm$ 100 years to this point-estimate. However, TPQ dates that do not account for the likely distance between site and sea when the site was in use limits the further inferential steps that can be taken. Furthermore, the more common practice of adding a constant error to a point-estimate does not take into account that the rate of shoreline displacement varies both spatially and temporally, and therefore introduces bias to the result on both the spatial and temporal scale, while also overestimating the precision of the method in most cases.

@@

The analysis in @roalkvam2023 found that site phases ^14^C-dated to before 2500 BCE have a likely elevation above sea-level that can be reasonably approximated by a gamma distribution with shape =  0.286 and scale = 20.833. This is incorporated in the procedure for shoreline dating used for this study (Figure \@ref(fig:shoredate)D). When performing this dating procedure, the gamma distribution is sequentially stepped through and transferred to the calendar scale by uniformly distributing the probability across the years in the range between the lower and upper limit of the interpolated displacement curve. This procedure thus gives the shoreline date of a site when accounting for the likely elevation of the site above sea-level when it was in use [see @roalkvam2023 for details]. Given that the shoreline displacement curves have no inversions and should therefore before commutative [cf. @weninger2015, 545], each shoreline date is normalised to sum to unity. To reduce the computational cost of the simulation procedures to follow below, the gamma distribution is here stepped through at increments of 0.1 m and the calendar scale is kept at a resolution of 5 years.

@@

The implications of different approaches to shoreline dating and summing the results are illustrated in Figure \@ref(fig:sumshore). In Figure \@ref(fig:sumshore)A  the sites to be shoreline dated are plotted according to their elevation. Due to the variable rates of shoreline displacement within the study region, the same elevation does not necessarily equate to the same date, and so this does not directly inform us about their temporal distribution. For example, as the trajectory for sea-level regression is more pronounced towards the north-east, the same elevation at a location to the south-west implies a younger date compared to one situated further north-east.

When the frequency distribution of shoreline dates have been investigated in the past, this has typically involved finding point-estimates of dates by either using the mean of displacement curves, or by taking mean of the TPQ dates for each site (as illustrated in Figure \@ref(fig:shoredate)C). These point-estimates are then aggregated in bins of 200 or 500 years [e.g. @breivik2014; @solheim2018; @fossum2020; @mjærum2022; @tallavaara2020]. This approach is illustrated in Figure \@ref(fig:sumshore)B. The SSPD resulting from using the probabilistic method of shoreline dating, given in Figure \@ref(fig:sumshore)C, demonstrates that accounting for the site-sea relationship gives a substantially different result compared to the more traditional approach, which is clearly indicated by the shifts in the positions of peaks and troughs in the distribution after c. 8000 BCE.

@@

In the same way is with an RSPD, the SSPD cannot be directly interpreted as reflecting the intensity of sites over time, due to the summation issues introduced above. Furthermore, similar to how characteristics of the calibration curve can introduce bias to the RSPD, the same is true for the SSPD, where the local trajectory of relative sea-level change effectively functions as the calibration curve for each site. As the shoreline displacement curves are interpolated to the sites based on their location along a south-west/north-east gradient, each site is effectively associated with a unique shoreline displacement curve, provided they are not located on exactly the same isobase. With analogy to the radiocarbon methodology, this would be equivalent to each radiocarbon date being associated with a unique calibration curve. As it would be computationally prohibitive to interpolate the shoreline displacement trajectory for each date to be simulated in the Monte Carlo procedure, one shoreline displacement curve was initially interpolated to the centre of each of a series of 2 km wide line segments running perpendicular to the shoreline gradient between the extremes of the distribution of sites. These intervals were assigned a weight based on how the density of observed sites is distributed among them (Figure \@ref(fig:incpolys)).

@@

```{r spdmc, echo = FALSE, fig.cap = "Monte Carlo simulation for shoreline and radiocarbon dates. A) Summed probability of shoreline dated sites (n = 921) compared to an exponential null model. B) SSPD compared to a logistic null model. C) SSPD compared to a uniform null model. D) Summed probability of calibrated radiocarbon dates (n = 310, bins = 134) compared to an exponential null model. E) RSPD compared to a logistic null model. F) RSPD compared to a uniform null model.", out.width = "500px", fig.align="center", regfloat = TRUE}
knitr::include_graphics(here("analysis/figures/mc.png"))
```
For the RSPD, the logistic model achieved a p-value of 0.108, and could therefore not be rejected. Based on the available data, this model represents the best explanation of the data. For the SSPD all of the standard models could be rejected.
